\section*{Acknowledgements}

I thank, first and foremost, my advisor, Jasmin Blanchette, who gave me enough rope to entangle myself, but also enough sage advice for disentangling.
If any parts of this thesis are well-written, they should be attributed to him.
I also thank the other members of my doctoral committee, TODO.
Equally important were my close collaborators, Asta Halkjær From and Xavier Généreux, who provided both the insights and, occasionally, the motivation to see the research in Chapters~\ref{chap:aesop} and~\ref{chap:forward} through.

I have greatly benefited from stimulating discussions with my colleagues at VU Amsterdam, LMU Munich, AWS and beyond (as well as their proofreading skills).
Thank you, in particular, to Anne Baanen, Alexander Bentkamp, Henrik Böving, Tanguy Bozec, Wouter Brozius, Martin Desharnais-Schäfer, Gabriel Ebner, Cedric Flamant, Wan Fokkink, Udaya Ghai, Massin Guerdi, Son Ho, Marc Huisinga, Lydia Kondylidou, Soonho Kong, Alex Kontorovich, Rob Lewis, Kenji Maillard, Luca Maio, Leonardo de Moura, Zhizhen Qin, Femke van Raamsdonk, Aloïs Rosset, Balazs Toth, Sebastian Ullrich, Petar Vukmirović, Yiming Xu.
Apologies to the many people I have probably forgotten.
I am also grateful to AWS and my manager, Kanna Shimizu, for letting me work on this thesis on company time, and to the Lean Focused Research Organisation, particularly Kim Morrison, for keeping Aesop up to date with frequent Lean releases.

Finally, I thank my partner, parents, grandparents, sister and wider family for their unwavering support.
I dedicate this thesis to my grandfather, who, at over 90 years of age, will not stop sending me mathematical notes.
